---
title: "S1_probability: Definitions and exercises"
shorttitle: "probability for CSHL"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{S1_probability: Definitions and exercises}
  %\VignetteEncoding{UTF-8}
output: 
  bookdown::html_document2:
    toc: yes
    toc_depth: 2
    number_sections: FALSE
pkgdown:
  as_is: true
---

```{r setup, echo=FALSE, results="hide"}
suppressMessages({
 suppressPackageStartupMessages({
  library(CSHstats)
  library(survival)
  library(ggplot2)
library(mvtnorm)
#  library(curatedTCGAData)
  })
 })
```

# Learning objectives

- Gain familiarity with concepts of simulation in R using sampling from
specified sets (e.g., a deck of cards) and using random number generation
    - understand the use of `sample()`
    - understand how to use `set.seed()` to control the behavior of random number generation
- Estimate event probabilities for specified experimental setups
    - distinguish the sample size of an experiment and the replication size of a simulation
- Explain the relationship between precision of a probability estimate and the sample
size of the associated experiment
- Define and use probability mass functions for discrete outcomes
    - for binomially distributed outcomes
    - for outcomes governed by the Poisson and Negative Binomial models
    - in general
- Understand how to simulate data under null (e.g., fair deck) and alternative (e.g.,
biased deck) conditions
- Understand the difference between finite, countably infinite, and uncountably
infinite sample spaces
- Define and use probability density functions and cumulative distribution functions
for continuous outcomes


# Overview

Our concept of probability is that of "long run relative frequency".

We'll work with several models of random events to make this more concrete.

# The probability of a 0-1 (binary or dichotomous) event

## Simulating shuffled playing cards

### Random permutations with `sample()`

We'll take for granted that `sample(x, size=length(x), replace=FALSE)` in R
achieves a goal of "shuffling" elements of `x`.  Thus we assume that,
if `x` is a vector in R, and
```
s1 = sample(x, size=length(x), replace=FALSE)
s2 = sample(x, size=length(x), replace=FALSE)
```
then no aspect of the ordering of elements in 
`s1` can be used to predict anything about the ordering of elements of `s2`.

In other words, `sample(x, size, replace=FALSE)` is taken as a primitive operation
that permutes the elements of `x` in a "random" way.

```{r demo1a}
set.seed(5432) # initialize, for reproducibility, the random number generator
sample(1:5, replace=FALSE) # permute (1,2,3,4,5)
sample(1:5, replace=FALSE) # a new, unpredictable, permutation
```

#### Exercises

**1: Why is the second permutation produced above termed "unpredictable"?**

**2: Predict the outcome of `sample(1:5, replace=FALSE)`, run just
after the one shown above.**

### Card deck definitions and operations

Our card deck is a vector with 52 elements.  Unicode characters encode the "suits".

If you have not already installed CSHstats (or did so a long time ago) use
```
BiocManager::install("vjcitn/CSHstats")  # install.packages("BiocManager") if necessary
```
to get access to the software needed to do the computations in this document.

```{r getli}
library(CSHstats)
d = build_deck()
d
unique(suits(d))
unique(faces(d))
table(suits(d), faces(d))
```

A fair deck has one card for each combination of "face" and "suit".

#### Exercises

**3: Write the code that produces a new deck that is
fair except that it has two copies
of the ten of hearts.  Verify that your new deck has this property.**

### Shuffling the deck

A reproducible shuffling event can be programmed as follows:

```{r doshuf}
set.seed(1234)  # any numeric seed will do but you need to record it
shuffle_deck = function(d) sample(d, size=length(d), replace=FALSE)
head(d)         # top 6 cards
head(shuffle_deck(d))
```

### The top card after a shuffle

```{r lktop}
set.seed(4141)
top_draw = function(d) shuffle_deck(d)[1]
top_draw(d)
top_draw(d)
```

#### Exercises 
**4: What is the top card in the next shuffle?**

**5: Write code to get the third card.**

**6: What is the probability that the top draw of a shuffled fair deck is a heart?**

## Estimating the probability of an event

### A simulation process: repeated shuffles and draws

Here's a simple way of estimating the probability
that the top draw of a shuffled fair deck is a heart, construed as long run frequency.

We'll simulate 100 shuffles and save the results of testing the suit of the top card.
```{r doh}
heart_sign = function() "\U2661"  # unicode U2661, prepend backslash
heart_sign()
res = rep(NA, 100)
for (i in 1:100) {
 res[i] = (suits(top_draw(d)) == heart_sign()) # TRUE if top card is a heart
}
head(res)
### relative frequency of the event "suit of top card is 'heart'"
sum(res)/length(res)  
```

A more concise approach with R is:
```{r do2}
mean(replicate( 100, suits(top_draw(d)) == heart_sign()) )
```

#### Exercises

**7: How do we interpret the difference between the two probability estimates given here?**

### Replicating the estimation process

Let's intensify our investigation with an aim
of understanding the uncertainty of estimation.

We'll define a variable that gives the size of our "experiment": we are shuffling
100 times and we'll refer to this as the sample size, `SSIZE`

```{r ssizedef}
SSIZE = 100
```

We will study the estimation procedure by replicating the experiment `N_REPLICATION` times.

```{r nrepdef}
N_REPLICATION = 500
```

```{r do3, eval=FALSE}
set.seed(10101) # initialize, for reproducibility, the random number generator
doubsim = replicate(N_REPLICATION, 
    mean(replicate( SSIZE, suits(top_draw(d)) == heart_sign()) )
    )
```
```{r do3real,echo=FALSE}
data(doubsim)
```
```{r lkhist1}
hist(doubsim, xlim=c(0,1))
```

#### Exercises

**8: How can we make the estimate of the probability of the event "suit of top card is 'heart'" more precise?**

Answer: increase `SSIZE`.

```{r do4, eval=FALSE}
SSIZE = 500
set.seed(101012)
doubsim2 = replicate(N_REPLICATION, 
    mean(replicate( SSIZE, suits(top_draw(d)) == heart_sign()) )
    )
```
```{r doret,echo=FALSE}
data(doubsim2)
```
```{r lkhj}
hist(doubsim2, xlim=c(0,1))
```

With an increased "sample size",
we have reduced variation in our experiment-to-experiment estimates of the probability 
of heart as suit of top card.

## Formal probability models

Intuitively, the probability of drawing a heart from a well-shuffled
fair deck is 1/4.  If we repeat the shuffling and drawing one hundred times,
we expect **around** 25 draws to reveal a heart.

Formal probability models enable us to reason systematically about
what we mean by **around** in our description of our expectation.
With these models we can also create accurate predictions of likely outcomes in
more complex events.

It is useful to define the **sample space** for a probability model,
which is the set of all possible outcomes of the random process being
modeled.  For our top-card-drawn example, the sample space 
is the set of suits, because we disregard the
face value.  For the fair deck, each point
in the sample space has probability $1/4$.
Events of interest are "top card is a heart"
and "top card is not a heart".  The probabilities of
these events are derivable from the probabilities of
the constituent outcomes.

### Using the binomial probability model

A series of independent dichotomous events (true or false, zero or one, heart or non-heart)
can be modeled using a **probability mass function** for the binomial distribution.  There are two parameters,
$p$ and $n$, where $p$ is the (unobservable) probability of the event (say "suit of top card is 'heart'")
and $n$ is the number of **independent** trials in which the random dichotomy is observed.  In $n$ "trials",
if the event has probability $p$, the probability of seeing the event $x$ times is
$$
Pr(X = x; n, p) = {\binom{n}{x}} p^x(1-p)^{n-x}
$$
where have written $X$ to denote the random quantity and $x$ to denote its realization.

So for a single draw, with $X$ the count of hearts seen in the draw, we have
$$
Pr(X=0; 1, 1/4) = 1-1/4 = 3/4
$$
$$
Pr(X=1; 1, 1/4) = 1/4
$$




#### Exercises

**9: Modify the production of `doubsim2` so that the $x$-axis for the
histogram has units "number of times the top card is a heart".**

```{r do5, eval=FALSE, echo=FALSE}
set.seed(43212)
doubsim3 = replicate(N_REPLICATION, 
    sum(replicate( SSIZE, suits(top_draw(d)) == heart_sign()) )
    )
```
```{r ret3,echo=FALSE}
data(doubsim3)
```
```{r dodou3h}
hist(doubsim3, xlim=c(.15*500,.35*500), xlab="count of draws with heart as suit of top card")
```

### Visualizing the model and the data

The formula given above tells us how frequently we will observe a given count.  The
R function `dbinom` can compute the probability, which we multiply by the number
of realizations to get the height of the histogram.

```{r sketch}
hist(doubsim3, xlim=c(.15*500,.35*500), 
   xlab="count of draws with heart as suit of top card", ylim=c(0,115))
points(80:160, 2500*dbinom(80:160, 500, 13/52), pch=19, cex=.5)
legend(78, 110, pch=19, legend="scaled dbinom(x, 500, .25)", bty="n", cex=.85)
```

Notice that the histogram, by virtue of its binning of the counts, does
not seem to reflect the shape of the theoretical frequency function given by the dots.
This can be remedied by increasing the number of replicates used.  

There is considerable research regarding the design of histogram displays.
See `?nclass.Sturges` for references.  One unpleasant feature of the
display for doubsim2 is that it seems to imply an asymmetric distribution.
Another is the way it "cuts off" at the extremes.

#### Exercises

**10: Write a couple of sentences carefully distinguishing the
displays of doubsim2 and doubsim3 above.  Explain the choice of
the heuristic name "doubsim".**

## A biased deck

Recall the layout of the fair deck:
```{r dobi}
table(suits(d), faces(d))
```
We will make a copy of one card and remove one:
```{r dobi2}
bd = d
bd[18] = bd[3]
table(suits(bd), faces(bd))
```

### Exercises

**11: What is the probability that the top card drawn after
a shuffle of `bd` is a heart?**

**12: What is the probability that the top card drawn after
a shuffle of `bd` is a club?**

# Probability models for categorical outcomes

Thus far we have focused on the dichotomy: top draw is heart or not.
We can consider all the possible suits as a 4-valued response.

## A contingency table

Let's simulate the process of drawing the top card after
shuffling, and tabulate the suits observed.

```{r dotab1}
set.seed(4321)
table( tops <- replicate(500, suits(top_draw(bd))) )
```

### Exercises

**13: What would you expect for the counts for suits in the table above
if the deck were fair?**

## Multinomial model and simulation

We adopted the binomial model for the number of top draws
with suit "heart" in a fixed number of shuffles:

$$
Pr(X = x; n, p) = {\binom{n}{x}} p^x(1-p)^{n-x}
$$

With the fair deck, we have $p= 1/4$.  A generalization
to a vector of responses is the multinomial model.  We
can use this for the (ordered) vector of counts of
top draws yielding different suits.  For this problem
we have parameters $N$ (number of trials), $k$ (number
of categories), and $p_1, \ldots, p_k$, the category-specific
probabilities.  The realizations are denoted $x_1, \ldots, x_k$
and we have $\sum_i x_i = N$.

Now the probability model is defined in terms of random
vectors and vectors of probabilities:

$$
Pr(X_1 = x_1, \ldots, X_k = x_k) = \frac{n!}{x_1! \cdots x_k!} p_1^{x_1} \cdots p_k^{x_k}
$$

This provides a more elegant way of producing frequency distributions
of the suit of the top draw.  Here we introduce pseudorandom number
generation from the multinomial model, using `rmultinom`.

```{r lkmn}
NREP = 10000
SSIZE = 500
mnmat = rmultinom(NREP, SSIZE, rep(.25,4))
rownames(mnmat) = c("\U2661", "\U2662", "\U2663", "\U2664")
mnmat[,1]  # one draw
apply(mnmat,1,mean)
```

Notice that we did not use the deck `d` in producing this matrix
of counts.  Previously we applied `sample()` to the 52-vector of
cards.  Now we use the model to develop the data
of interest.

### Exercise

**14: Modify the call to `rmultinom` to obtain distributions of
top card suits for the biased deck `bd`.  
Hint: Change the part of the call involving `rep()`.**

## The Poisson distribution

### Modeling collections of counts

Outcomes taking the form of integer counts arise in many
appications.  As an example, a quarry floor is divided into
squares about one meter on the side.  30 squares were searched
for a fossil and the number of specimens per square is recorded.
No square had 5 or more specimens.  (Example from _Probability
Models and Applications_, I. Olkin, L. Gleser, C. Derman, ch 6.3.)

The table below includes `nspec`, the number of specimens in
a square, `freq`, the number of squares having the corresponding
number of specimens, and `pred`, a predicted number of specimens
based on the Poisson model with mean parameter 0.73 specimens
per square.
```{r lkfoss}
foss = data.frame(nspec=0:4, freq=c(16,9,3,1,1), pred=round(30*dpois(0:4, 0.73), 2))
foss
```

To derive the quantities `pred`, we used `dpois` for the counts per square,
with mean parameter set at 0.73.  To derive this, we use 
```{r lksum}
sum(foss$nspec * foss$freq)/30
```
to obtain the average number of fossils per square.  The Poisson model with
mean $\lambda$ is
$$
Pr(X = x; \lambda) = \frac{e^{-\lambda}\lambda^k}{k!}
$$

### Exercise

**15: Use the formula above to explain the prediction of 14.46 squares with zero fossils.**

### A countably infinite sample space

The binomial and multinomial models discussed above have finite
discrete sample spaces.  The sample space for the Poisson model 
is the set of non-negative integers.  The mass function arises
from the fact that $e^\lambda = \sum \lambda^k/k!$, the sum
taken over all non-negative integers.

## The negative binomial model

The mass function is
$$
p(x; \theta, \mu) = \frac{\Gamma(\theta + x)}{\Gamma(\theta)y!} 
   \frac{\mu^y \theta^\theta}{(\mu + \theta)^{\theta + y}}
$$

This will be useful for later material.  


## Mean, variance

The sample mean for the vector $x = (x_1, \ldots, x_n)$
is defined as $\bar{x} = n^{-1}\sum_i x_i$.  The (unbiased estimator
of the) sample variance
is $(n-1)^{-1}\sum_i (x_i = \bar{x})^2$.

We will sometimes refer to mean and
variance in the context of probability models
rather than samples.  The mean for a continuous
distribution is defined as

For a discrete distribution with probability
mass function $p$ the mean is

$$
E(x) = \sum x p(x),
$$
The sum here is taken over the sample space
for the associated model.

Writing $\mu$ for the mean value of the distribution
under study, the variance of a discrete distribution
is 
$$
V(x) = \sum (x-\mu)^2 p(x) dx.
$$


# Models for continuous outcomes

The models we've considered thus far are for discrete
responses -- the sample space is either finite or countably infinite.

For continuous responses, the sample space is uncountably
infinite.  The sample space may be an interval of the real line,
or the whole real line.  We will discuss how to
define probabilities for subsets of the sample space.

## Univariate response

For concreteness, we will use RNA-seq data derived
from The Cancer Genome Atlas.  We'll focus on `r {data(yy1_ex); length(yy1_ex)}`
values of normalized RSEM-based estimates of
expression of YY1 in adrenocortical carcinoma.


```{r lkacc1, eval=FALSE, echo=FALSE}
accex = curatedTCGAData("ACC", "RNASeq2GeneNorm",version="2.0.1", dry.run=FALSE)
```
```{r getdacc, echo=FALSE, echo=FALSE}
data(accex)
```
```{r lkaccex, echo=FALSE, eval=FALSE}
accex
accrna = experiments(accex)[[1]]
yy1_ex = assay(accrna["YY1",])
class(yy1_ex)
```
```{r lkacc2}
data(yy1_ex)
head(yy1_ex)
```

We will refer to this as a vector of "real valued" outcomes.


### Cumulative distribution function; quantiles

We'll use capital letters to denote random variables.
The formal definition of "random variable" is given
in textbooks; we will use the term informally to
refer to quantities we regard as random.

The function $F(x) = \mbox{Pr}(X < x)$ is called
the cumulative distribution function (CDF) for 
the random variable $X$.  For the
YY1 measures, we can produce a visualization of
the associated CDF:

```{r lkecdf}
plot(ecdf(yy1_ex))
abline(h=.8, lty=2)
arrows(1583, .8, 1583, 0, lty=2)
```

The quantiles of a distribution are given by
the inverse of the CDF.  The quantile function
takes a number $p$ between 0 and 1 and returns
the number $F^{-1}(p)$.

The graph above shows that the 0.80 quantile
of the distribution of our YY1 vector is about 1584.

### Exercise

**16:  Show that the proportion of YY1 observations 
exceeding 1584 is approximately 20%.**

### Histogram, probability density function

The histogram is a tunable display of the relative
frequencies of values in a vector.

```{r lkhi}
par(mfrow=c(2,1), mar=c(4,3,1,1))
hist(yy1_ex, xlim=c(400, 2600))
hist(yy1_ex, breaks=20, xlim=c(400, 2600))
```

The probability density function for a
continuous random variable $X$ satisfies

$$
\int_a^b f(x) dx = \mbox{Pr}(a < X < b)
$$

We can estimate the density function for a
sample in various ways.  Here is a simple
illustration:

```{r lkden}
plot(dd <- density(yy1_ex))
```

It is of some interest to use `approxfun` and numerical
integration with the density estimate given above,
to estimate the probability that YY1
expression lies in a given interval.

```{r doapprox}
dfun = approxfun(dd$x, dd$y)
integrate(dfun, 1000, 1500)  # density-based
mean(yy1_ex >=1000 & yy1_ex <= 1500) # empirical
```

Is this acceptable?  Could changing the bandwidth for the
density estimator produce better results?  Is there
a potential overfitting problem?




### Some widely used models for continuous responses

We'll use simulation to illustrate the shapes of
various distributions.  R makes this very simple with
a family of functions whose names begin with `r`.

#### Uniform distribution on [0,1]

The density function is $f(x) = 1$  if $x \in [0,1]$ and 0 otherwise.

```{r lkunif}
hist(runif(10000, 0, 1), prob=TRUE)
```

##### Exercise

**17:  Use code like that above to obtain
the value of the density of a random
variable with uniform density on the interval [0, 2].**

#### Gaussian model

The Gaussian distribution is also called "normal".
Its shape and position on the real line are determined
by its mean and variance.  Symbolically, 
the model is often written $N(\mu, \sigma^2)$, and
the density function is 
$f(x) = 1/\sqrt{2\pi \sigma^2} \exp\{(x-\mu)^2/2 \sigma^2\}$.

```{r lknorm}
hist(rnorm(10000, 0, 1), prob=TRUE)
lines(seq(-4,4,len=100), dnorm(seq(-4,4,len=100)))
```

#### Exponential and Gamma models for survival data

The exponential model (not to be confused with
exponential family models) is defined for random
variables with positive values.  An example
is survival times.  We'll use a dataset from the survival
package to illustrate.

```{r lkmd}
library(survival) # use myeloma records from Mayo clinic, remove censored times
md = myeloma$futime[myeloma$death==1 & myeloma$entry == 0]
hist(md, breaks=20, prob=TRUE, ylim=c(0,.001))
mean(md)
ss = seq(0,8000,.1)
lines(ss, dexp(ss, rate=1/mean(md)))
```

The density function for the exponential model is $f(x) = \lambda e ^ {-\lambda x}$.
In this formulation, $\lambda$ is referred to as the rate parameter.  The
mean of the distribution is then $1/\lambda$.

##### Exercise

**18:  How close is the median of the exponential model
with rate 1/1045 to the sample median of the myeloma
survival times?  How does this relate to the interpretation of
the plot above?**

## Mean and variance for continuous models

For continuous distributions, the mean value is
$$
E(x) = \int x f(x) dx.
$$
Writing $\mu$ for the mean value of the distribution
under study, the variance of a continuous distribution
is 
$$
V(x) = \int (x-\mu)^2 f(x) dx.
$$



## Multivariate response

We've seen that the use of univariate probability models has
many facets.
The concept of *joint distribution* of *multiple* random quantities
is central to reasoning about interactions among components
of complicated processes.

In the following example, we have obtained normalized expression measures 
for EGR1 and FOS in adrenocortical carcinoma tumors studied
in TCGA.  We'll use built-in bivariate density estimation to
show the relationship between expression of these two
transcription factors in ACC.

```{r lkcont}
data(fos_ex)
data(egr1_ex)
bivdf = data.frame(fos_ex, egr1_ex)
library(ggplot2)
ggplot(bivdf, aes(x=log(fos_ex), y=log(egr1_ex))) + geom_point() + geom_density_2d()
```

### Joint cumulative distribution function

We'll focus on the case of continuous bivariate response denoted $(X,Y)$.  The joint cumulative
distribution function (cdf) is
$$
F(x,y) = Pr(X < x, Y < y)
$$
So, for log FOS and log EGR1, we can evaluate $F(7,7)$ as follows:
```{r lkjcdf}
mean(log(fos_ex)<7 & log(egr1_ex)<7)
```

### Covariance matrix, bivariate case

The covariance between two random variables is $Cov(X,Y) = EXY - (EX)(EY)$;
the covariance matrix is symmetric, has the variances on the diagonal,
and the covariance off the diagonal.
```{r lkcov}
var(log(fos_ex)) # univariate
covmat = var(cbind(log(fos_ex), log(egr1_ex))) # matrix
covmat
```

### Using the bivariate normal model

Given estimates of the bivariate mean and covariance, we can use
the multivariate normal cumulative density function to produce
contours of the distribution that will have elliptical shapes,
as opposed to the wiggly contours we saw above.

```{r getdf}
bmean = c(mean(log(fos_ex)), mean(log(egr1_ex)))
library(mvtnorm)
xgrid = seq(3,10,.05)
ygrid = seq(3,10,.05)
ngrid = length(xgrid)
nd = matrix(NA, ngrid, ngrid)
for (i in 1:ngrid) {  # inefficient!
 for (j in 1:ngrid) {
  nd[i,j] = dmvnorm(c(xgrid[i],ygrid[j]), bmean, covmat)
  }
 }
contour(xgrid, ygrid, nd, ylab="log EGR1", xlab="log FOS")
points(log(fos_ex), log(egr1_ex), col="gray", pch=19)
```

The estimate of $F(7,7)$ using this model is
```{r do77}
pmvnorm(upper=c(7,7), mean=bmean, sigma=covmat)
```


### Probabilistic independence and dependence

### Measures of correlation

