---
title: "S4_inference: Definitions and exercises"
shorttitle: "inference for CSHL"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{S4_inference: Definitions and exercises}
  %\VignetteEncoding{UTF-8}
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 2
    number_sections: FALSE
pkgdown:
  as_is: true
---

# Inference

## The fairness hypothesis

Let $C_1$ denote the suit of the top card revealed after a fair shuffle.
We can state a hypothesis about the fairness of the deck
under repeated draws of top card after shuffling as

```{r defs, echo=FALSE}
spade_sign = function() "\U2660"
diamond_sign = function() "\U2662"
club_sign = function() "\U2663"
```

$$
H_0: Pr(C_1 = \heartsuit) = Pr(C_1 = \diamondsuit) = Pr(C_1 = \clubsuit) = Pr(C_1 = \spadesuit) = 1/4
$$

## Operating characteristics: Type I and Type II errors

In a frequentist framework for statistical inference, we
define procedures for testing (null) hypotheses with specified
error probabilities.  

- A **Type I error** occurs when the
null hypothesis is true but the test results in the assertion that it is
false.  Traditionally we try to keep the probability of Type I errors
below 5\%.

- A **Type II error** occurs when the
null hypothesis is false but the test does not result in an
assertion that it is false.  Traditionally we try to keep the
probability of Type II errors below 20\%.
 
- Synonyms: We often use **size** as synonymous with Type I error probability,
and **power** as one minus the Type II error probability

### Exercise

**1: Propose a procedure for testing $H_0: Pr(C_1 = \heartsuit) = 1/4$.  Assume you have
the results of top card draws from N shuffles.  What would such a procedure look like?**

## Simulating a series of experiments; parameter estimate

```{r dosim}
suppressPackageStartupMessages({
suppressMessages({
library(CSHstats)
library(EnvStats)
})
})
d = build_deck()
shuffle_deck = function(d) sample(d, size=length(d), replace=FALSE)
heart_sign = function() "\U2661"
set.seed(4141)
top_draw = function(d) shuffle_deck(d)[1]
N = 100
mydat = replicate(N, suits(top_draw(d))==heart_sign())
phat = function(dat) sum(dat)/length(dat)
phat(mydat)
```
With 100 shuffles we have an estimate of the probability of a
heart.  Is our experimental result consistent with $H_0$?

With a larger sample size:
```{r dolar}
N = 1000
mydat = replicate(N, suits(top_draw(d))==heart_sign())
sum(mydat)/length(mydat)
```

## An improvised test

Let's use the procedure `|phat(dat)-.25|>.01` as our criterion for
rejecting
$H_0: Pr(C_1 = \heartsuit) = 1/4$
What is the Type I error rate
for a fair deck for the experiment based on 100 shuffles?

```{r chknull, cache=TRUE}
N = 100
NSIM = 1000
tsts = replicate(NSIM,
   abs(mean(replicate(N, suits(top_draw(d))==heart_sign()))-.25)>.01)
mean(tsts)
```

Parameterize the "delta", which was 0.01 in the previous run.  Let's
get the rejection rate estimates for two sample sizes:

```{r lkdel,cache=TRUE}
prej = function(delta=.01, nullval=.25, NSIM, N) {
 mean(replicate(NSIM,
   abs(mean(replicate(N, suits(top_draw(d))==heart_sign()))-nullval)>delta))
}
prej(delta=.02, NSIM=1000, N=100)
prej(delta=.02, NSIM=1000, N=500)
```

Our home-cooked test has a Type I error rate that is much too high
for standard practice, and depends on the sample size.

## A properly calibrated procedure

Let's instead use the built in procedure for testing hypotheses
with binomial 
outcomes.  

The `brej` function defined below has two layers:
- inner: given N draws, test that the sum of the number of
hearts is consistent with a specified null value (1/4 by
default), with a specified Type I error rate (argument alpha)
- outer: replicate the inner process NSIM times to obtain
NSIM indicator variables taking value 1 if the test rejected
and 0 otherwise

We then estimate rejection probabilities by taking the means
of the indicator variables for different experimental and
replication setups.

```{r tryz, cache=TRUE}
brej = function(deck, nullval=.25, alpha=0.05, NSIM, N) {
 replicate(NSIM, {
   dat = replicate(N, suits(top_draw(deck))==heart_sign())
   binom.test(sum(dat), N, nullval)$p.value < alpha
   })
}
mean(brej(d,NSIM=1000, N=100))
mean(brej(d,NSIM=1000, N=500))
mean(brej(d,NSIM=1000, N=100, alpha=0.01))
```

We have an indication here that the procedure stabilizes the Type I error
rate for different designs (sample sizes) and can accommodate different
significance levels.

# Power curve

It is often useful to sketch the power of a test procedure
over a series of values of an element of the experimental/testing
setup.  We'll consider how the power to reject the null
varies with the sample size, for two alternatives to the null.

We make a biased deck by switching one diamond to a heart.
For such a deck, the probability of a heart is 14/52, greater
than 1/4.

```{r dobi}
bd = d
bd[18] = bd[4]
table(suits(bd), faces(bd))
```
What kind of sample size do we need to get good power to detect this
exception?

```{r docur, cache=TRUE}
ssizes = c(10,50, 100, 300,750)
rejps = sapply(ssizes, function(x) mean(brej(bd, NSIM=100, N=x)))
plot(ssizes, rejps, main="detect one extra heart")
```

With two switches, the probability of a heart increases to 15/52.

```{r domor, cache=TRUE}
bd[19] = bd[5]
table(suits(bd), faces(bd))
rejps = sapply(ssizes, function(x) mean(brej(bd, NSIM=100, N=x)))
plot(ssizes, rejps, main="detect two extra hearts")
rejps_n = sapply(ssizes, function(x) mean(brej(d, NSIM=100, N=x)))
plot(ssizes, rejps_n, main="preserve Type I error rate")
```

# Conclusions

- Classical frequentist inference aims to control probabilities of
errors of Type I (reject a null hypothesis that is true) and Type II
(fail to reject a false null).  The power of a test procedure is
one minus its Type II error probability.

- Typically the null hypothesis specifies the value of a parameter
of a population distribution 
    - in the case of the card deck, the
hypothesis of fairness of deck and shuffling procedure entails
the probability of drawing a heart after a shuffle is 1/4 
    - the experimental setup entails that the probability distribution
of the sum of the number of hearts drawn as top card under N shuffling events
is binomial(N,1/4)

- R's `binom.test` is given 
    - the sum of the number of hearts observed,
    - the number of trials, 
    - the parameter value defining the
 null hypothesis, 
    - the desired type I error rate,

and produces a
critical region using the binomial distribution

When the sum falls in the critical region, the null hypothesis is rejected

- Verification of achievement of the desired Type I and Type II error
probabilities for a testing procedure can be pursued via simulation

- The power curve for a sequence of experimental conditions can be produced
via simulation

- Analytical tools for power computation are also available.  For example,
in the EnvStats package, we find
```{r dopow}
library(EnvStats)
propTestPower(750, p.or.p1=13/52, p0=.25 )
propTestPower(750, p.or.p1=14/52, p0=.25 )
propTestPower(750, p.or.p1=15/52, p0=.25 )
```
An exercise is to reconcile these results with the simulation-based
results given above.

Another is to find the sample size needed to achieve 80% power to detect
a single switch from diamond to heart in an otherwise fair deck.
