---
title: "S5b: hypothesis testing in nonlinear regression"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{S5b: hypothesis testing in nonlinear regression}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Logistic curve with response decreasing in x

Let a denote the asymptotic response at large values of x
and b denote the response at x=0.  c will denote a
slope-like parameter, and d the midpoint of transition
between asymptotes.

```{r lklo}
a = 4
b = 1
c = 0.8
d = 4.5
x = 0:10
y = a + (b-a)/(1+exp(-c*(x-d)))
plot(x,y,type="b", ylim=c(0,5))
```

A variety of S-shaped curves can be produced depending
on choices of the parameters.

Exercise: Increase the value of the parameter c, recompute
y, and overlay on the plot with lines() and lty=2.

# Fitting the model to data

The `nls` function can fit this model.  We have some data
that were crafted to illustrate a "pair" of curves
that can be regarded as shifted instances of common shape.
Here the color distinguishes the elements of the pair.

```{r lkdat, message=FALSE}
library(CSHstats)
data(nonlindat)
plot(y~x,data=nonlindat, col=nonlindat$z+1, pch=19)
```

We are modeling a pair of experiments in which the
red points arise from a different condition than
the black points.  The null hypothesis that the curves
are identical is tested using a parameter 'del'
measuring the shift in the response curve.
An indicator variable z takes value 0 for control
condition and 1 for the active treatment.

```{r donls}
nl1 <- nls(y ~ 
 a + ((b - a)/(1 + exp(-c * (x - d*(1+del*z))))), 
 start = list(a = min(y), b = max(y), c = 1, 
 d = round(median(x)), del=0), data=nonlindat)
summary(nl1)
```

Let's use the model to overlay predicted response curves
on the original data.

```{r relook}
plot(y~x,data=nonlindat, col=nonlindat$z+1, pch=19)
nd = data.frame(x=1:8,z=0)
y1 = predict(nl1, newdata=nd)
lines(1:8, y1, lty=2)
nd2 = data.frame(x=1:8,z=1)
y2 = predict(nl1, newdata=nd2)
lines(1:8, y2, lty=2, col="red")
```

In summary, there is some evidence of a shift, with a p-value of 0.073.

Exercise (advanced): Test the hypothesis that the two conditions
have different asymptotic values as x increases.
