---
title: "S7 Neural net concepts and applications"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{S7 Neural net concepts and applications}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setuppp, echo=FALSE,results="hide"}
suppressPackageStartupMessages({
suppressMessages({
library(littleDeep)  # vjcitn/littleDeep, includes python interface
library(DT)
})
})
```
# Introductory remarks

- learning a nonlinear regression model and putting it to use
- "layers" of a multilayer neural net and their correspondence to
patient characteristics (molecular, clinical) and treatment outcomes
    - input, output, loss function
- [excellent high level overview](https://stats.stackexchange.com/questions/114385/what-is-the-difference-between-convolutional-neural-networks-restricted-boltzma) comparing convolutional neural nets and autoencoders

# Road map

- ImageArray
- [ISLR](https://hastie.su.domains/ISLR2/ISLRv2_corrected_June_2023.pdf)
(also a [python version](https://hastie.su.domains/ISLP/ISLP_website.pdf) Convolutional Neural Net example (also see [Jeremy Jordan](https://www.jeremyjordan.me/convolutional-neural-networks/)

- Sfaira model zoo of embeddings and classifiers

# Images and categories

```{r librar}
library(littleDeep)
data(ciftrain1k)
ciftrain1k
n <- preview(ciftrain1k)
```

We can filter an ImageArray
```{r lkty}
preview(filterByType(ciftrain1k, "orange"))
```

Question on bias: Does the machine recognize the orange or just the "more orange" color distribution?
How do you "normalize" images so that the "object itself" presents only the "essential features"?

# A trained classifier that ingests JPGs as arrays and returns scores for resemblance to
pre-trained categories

Check [here](https://vjcitn.github.io/littleDeep/articles/A1_three_cl.html)

I'll demonstrate 'jpeg_shrinker' which could be installed on intel macs but not on M1 macs at
this time.  This program will convert a jpeg image to 32x32 resolution and use a trained
CNN to classify the content.

```{r lkcode}
islr_cnn
```

# Sfaira: a model zoo of pretrained embedders and cell-type classifiers from Theis Lab

We'll look at [the pkgdown site](https://vjcitn.github.io/BiocSfaira), specifically "get started"
tab.
